A multivariate skew-normal finite mixture model for analysis of infant development trajectories

### 2019-05-23 manuscript/manuscript.tex
appendix

%  ALWAYS USE THE referee OPTION WITH PAPERS SUBMITTED TO BIOMETRICS!!!
%  You can see what your paper would look like typeset by removing
%  the referee option.  Because the typeset version will be in two
%  columns, however, some of your equations may be too long. DO NOT
%  use the \longequation option discussed in the user guide!!!  This option
%  is reserved ONLY for equations that are impossible to split across 
%  multiple lines; e.g., a very wide matrix.  Instead, type your equations 
%  so that they stay in one column and are split across several lines, 
%  as are almost all equations in the journal.  Use a recent version of the
%  journal as a guide. 
%  
\documentclass[useAMS,referee]{biom}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{blindtext}
\usepackage{booktabs}

@@

The multivariate skew-normal regression model can be written as follows.

$$\mathbf{Y} = \mathbf{X}\mathbf{B} + \mathbf{t} \boldsymbol\psi^T + \mathbf{E} = \mathbf{X}^* \mathbf{B}^* + \mathbf{E}$$

@@

The conditional likelihood for this model is given below.

$$p(\mathbf{Y}|\mathbf{X}^*,\mathbf{B}^*,\boldsymbol\Sigma) \propto |\boldsymbol\Sigma|^{-n/2}\exp \left \{ -\frac{1}{2} \tr(\mathbf{Y} - \mathbf{X}^* \mathbf{B}^*)^T(\mathbf{Y} - \mathbf{X}^* \mathbf{B}^*)\boldsymbol\Sigma^{-1} \right \}$$

@@

For a given observation vector $\mathbf{y} \sim N_p(\boldsymbol\mu, \boldsymbol\Sigma)$, we allow for missingness in at most $p - 1$ of the multivariate outcomes through the use of a conditional imputation step embedded within our Gibbs sampler. Suppose $\mathbf{y}$ contains $q$ missing observations and can be partitioned into two vectors $\mathbf{y_{1}}$ and $\mathbf{y_{2}}$ such that $\mathbf{y_{1}}$ is a $q \times 1$ vector of missing observations and $\mathbf{y_{2}}$ is a $(p-q) \times 1$ vector of complete observations. Similarly, partition $\boldsymbol\mu$ and $\boldsymbol\Sigma$ as follows.

$$\boldsymbol\mu = \begin{bmatrix} \boldsymbol\mu_1 \\ \boldsymbol\mu_2 \end{bmatrix}$$

$$\boldsymbol\Sigma = \begin{bmatrix} \boldsymbol\Sigma_{11} & \boldsymbol\Sigma_{12} \\ \boldsymbol\Sigma_{21} & \boldsymbol\Sigma_{22} \end{bmatrix}$$

We will use these quantities to derive the conditional distribution $f(\mathbf{y_1}|\mathbf{y_2},\boldsymbol\mu,\boldsymbol\Sigma)$.

\begin{align*} 
f(\mathbf{y_1}|\mathbf{y_2},\boldsymbol\mu,\boldsymbol\Sigma) & \propto  f(\mathbf{y_1},\mathbf{y_2}|\boldsymbol\mu,\boldsymbol\Sigma) \\
& \propto \exp \left \{ -\frac{1}{2}(\mathbf{y} - \boldsymbol\mu)^T \boldsymbol\Sigma^{-1} (\mathbf{y} - \boldsymbol\mu) \right \} \\
& = \exp \left \{ -\frac{1}{2}\begin{bmatrix} \mathbf{y}_1 - \boldsymbol\mu_1 \\  \mathbf{y}_2 - \boldsymbol\mu_2 \end{bmatrix}^T \boldsymbol\Sigma^{-1} \begin{bmatrix} \mathbf{y}_1 - \boldsymbol\mu_1 \\  \mathbf{y}_2 - \boldsymbol\mu_2 \end{bmatrix} \right \} \\
& =  \exp \left \{ -\frac{1}{2}\begin{bmatrix} \mathbf{y}_1 - \boldsymbol\mu_1 \\  \mathbf{y}_2 - \boldsymbol\mu_2 \end{bmatrix}^T \begin{bmatrix} \boldsymbol\Sigma_{11} & \boldsymbol\Sigma_{12} \\ \boldsymbol\Sigma_{21} & \boldsymbol\Sigma_{22} \end{bmatrix}^{-1} \begin{bmatrix} \mathbf{y}_1 - \boldsymbol\mu_1 \\  \mathbf{y}_2 - \boldsymbol\mu_2 \end{bmatrix} \right \} \\
& = \exp \left \{ -\frac{1}{2}\begin{bmatrix} \mathbf{y}_1 - \boldsymbol\mu_1 \\  \mathbf{y}_2 - \boldsymbol\mu_2 \end{bmatrix}^T \begin{bmatrix} \boldsymbol\Sigma_{11}^* & \boldsymbol\Sigma_{12}^* \\ \boldsymbol\Sigma_{21}^* & \boldsymbol\Sigma_{22}^* \end{bmatrix} \begin{bmatrix} \mathbf{y}_1 - \boldsymbol\mu_1 \\  \mathbf{y}_2 - \boldsymbol\mu_2 \end{bmatrix} \right \}
\end{align*}

where the blockwise inversion formula is used to invert $\boldsymbol\Sigma$ as follows.

\begin{align*}
\boldsymbol\Sigma_{11}^* & = \boldsymbol\Sigma_{11}^{-1} + \boldsymbol\Sigma_{11}^{-1}\boldsymbol\Sigma_{12}(\boldsymbol\Sigma_{22} - \boldsymbol\Sigma_{21}\boldsymbol\Sigma_{11}^{-1}\boldsymbol\Sigma_{12})^{-1}\boldsymbol\Sigma_{21}\boldsymbol\Sigma_{11}^{-1}
\end{align*}

@@

\begin{thebibliography}{}

\bibitem{ } Cox, D. R. (1972). Regression models and life tables (with
discussion).  \textit{Journal of the Royal Statistical Society, Series B}
\textbf{34,} 187--200.

\bibitem{ }  Hastie, T., Tibshirani, R., and Friedman, J. (2001). \textit{The 
Elements of Statistical Learning: Data Mining, Inference, and Prediction}.
New York: Springer.

\end{thebibliography}

>>>>

%  ALWAYS USE THE referee OPTION WITH PAPERS SUBMITTED TO BIOMETRICS!!!
%  You can see what your paper would look like typeset by removing
%  the referee option.  Because the typeset version will be in two
%  columns, however, some of your equations may be too long. DO NOT
%  use the \longequation option discussed in the user guide!!!  This option
%  is reserved ONLY for equations that are impossible to split across 
%  multiple lines; e.g., a very wide matrix.  Instead, type your equations 
%  so that they stay in one column and are split across several lines, 
%  as are almost all equations in the journal.  Use a recent version of the
%  journal as a guide. 
%  
\documentclass[useAMS,referee]{biom}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{blindtext}
\usepackage{booktabs}
\usepackage{breqn}

@@

The multivariate skew-normal regression model can be written as follows. 
$$\mathbf{Y} = \mathbf{X}\mathbf{B} + \mathbf{t} \boldsymbol\psi^T + \mathbf{E} = \mathbf{X}^* \mathbf{B}^* + \mathbf{E}$$

@@

The conditional likelihood for this model is given below. 
$$p(\mathbf{Y}|\mathbf{X}^*,\mathbf{B}^*,\boldsymbol\Sigma) \propto |\boldsymbol\Sigma|^{-n/2}\exp \left \{ -\frac{1}{2} \tr(\mathbf{Y} - \mathbf{X}^* \mathbf{B}^*)^T(\mathbf{Y} - \mathbf{X}^* \mathbf{B}^*)\boldsymbol\Sigma^{-1} \right \}$$

We choose conjugate priors for $\mathbf{B}^*$ and $\boldsymbol\Sigma$ as follows. 
$$\boldsymbol\Sigma \sim \text{inverse-Wishart}(\mathbf{V}_0,\nu_0) $$
$$\mathbf{B}^*|\boldsymbol\Sigma \sim MN_{(m+1)\times p}(\mathbf{B}_0^*,\mathbf{L}_0^{-1},\boldsymbol\Sigma)$$

We now dervive the joint posterior distribution of the parameters $\mathbf{B}^*$ and $\boldsymbol\Sigma$.
\begin{align*} 
p(\mathbf{B}^*,\boldsymbol\Sigma|\mathbf{X}^*,\mathbf{Y}) & \propto  p(\mathbf{Y}|\mathbf{X}^*,\mathbf{B}^*,\boldsymbol\Sigma)p(\mathbf{B}^*|\boldsymbol\Sigma)p(\boldsymbol\Sigma)\\
&  \propto |\boldsymbol\Sigma|^{-n/2}\exp \left \{ -\frac{1}{2} \tr((\mathbf{Y} - \mathbf{X^*}\mathbf{B}^*)^T(\mathbf{Y} - \mathbf{X^*}\mathbf{B}^*)\boldsymbol\Sigma^{-1})\right \} \\
& \times |\boldsymbol\Sigma|^{-m/2} \exp \left \{ -\frac{1}{2} \tr((\mathbf{B}^* - \mathbf{B}^*_0)^T \mathbf{L}_0(\mathbf{B}^* - \mathbf{B}^*_0)\boldsymbol\Sigma^{-1})\right \} \\
& \times |\boldsymbol\Sigma|^{(\nu_0 + p + 1)/2} \exp \left \{ -\frac{1}{2} \tr (\mathbf{V}_0 \boldsymbol\Sigma^{-1}) \right \}
\end{align*}

@@

For a given observation vector $\mathbf{y} \sim N_p(\boldsymbol\mu, \boldsymbol\Sigma)$, we allow for missingness in at most $p - 1$ of the multivariate outcomes through the use of a conditional imputation step embedded within our Gibbs sampler. Suppose $\mathbf{y}$ contains $q$ missing observations and can be partitioned into two vectors $\mathbf{y_{1}}$ and $\mathbf{y_{2}}$ such that $\mathbf{y_{1}}$ is a $q \times 1$ vector of missing observations and $\mathbf{y_{2}}$ is a $(p-q) \times 1$ vector of complete observations. Similarly, partition $\boldsymbol\mu$ and $\boldsymbol\Sigma$ as follows.
$$\boldsymbol\mu = \begin{bmatrix} \boldsymbol\mu_1 \\ \boldsymbol\mu_2 \end{bmatrix} \ \ \ \ \ \boldsymbol\Sigma = \begin{bmatrix} \boldsymbol\Sigma_{11} & \boldsymbol\Sigma_{12} \\ \boldsymbol\Sigma_{21} & \boldsymbol\Sigma_{22} \end{bmatrix}$$

We will use these quantities to derive the conditional distribution $f(\mathbf{y_1}|\mathbf{y_2},\boldsymbol\mu,\boldsymbol\Sigma)$. 
\begin{align*} 
f(\mathbf{y_1}|\mathbf{y_2},\boldsymbol\mu,\boldsymbol\Sigma) & \propto  f(\mathbf{y_1},\mathbf{y_2}|\boldsymbol\mu,\boldsymbol\Sigma) \\
& \propto \exp \left \{ -\frac{1}{2}(\mathbf{y} - \boldsymbol\mu)^T \boldsymbol\Sigma^{-1} (\mathbf{y} - \boldsymbol\mu) \right \} \\
& = \exp \left \{ -\frac{1}{2}\begin{bmatrix} \mathbf{y}_1 - \boldsymbol\mu_1 \\  \mathbf{y}_2 - \boldsymbol\mu_2 \end{bmatrix}^T \boldsymbol\Sigma^{-1} \begin{bmatrix} \mathbf{y}_1 - \boldsymbol\mu_1 \\  \mathbf{y}_2 - \boldsymbol\mu_2 \end{bmatrix} \right \} \\
& =  \exp \left \{ -\frac{1}{2}\begin{bmatrix} \mathbf{y}_1 - \boldsymbol\mu_1 \\  \mathbf{y}_2 - \boldsymbol\mu_2 \end{bmatrix}^T \begin{bmatrix} \boldsymbol\Sigma_{11} & \boldsymbol\Sigma_{12} \\ \boldsymbol\Sigma_{21} & \boldsymbol\Sigma_{22} \end{bmatrix}^{-1} \begin{bmatrix} \mathbf{y}_1 - \boldsymbol\mu_1 \\  \mathbf{y}_2 - \boldsymbol\mu_2 \end{bmatrix} \right \} \\
& = \exp \left \{ -\frac{1}{2}\begin{bmatrix} \mathbf{y}_1 - \boldsymbol\mu_1 \\  \mathbf{y}_2 - \boldsymbol\mu_2 \end{bmatrix}^T \begin{bmatrix} \boldsymbol\Sigma_{11}^* & \boldsymbol\Sigma_{12}^* \\ \boldsymbol\Sigma_{21}^* & \boldsymbol\Sigma_{22}^* \end{bmatrix} \begin{bmatrix} \mathbf{y}_1 - \boldsymbol\mu_1 \\  \mathbf{y}_2 - \boldsymbol\mu_2 \end{bmatrix} \right \} \\
& = \exp \left \{ -\frac{1}{2} \left[ (\mathbf{y}_1 - \boldsymbol\mu_{cond})^T \boldsymbol\Sigma_{cond}^{-1}(\mathbf{y}_1 - \boldsymbol\mu_{cond})\right]\right \} 
\end{align*}
$$\Rightarrow \mathbf{y_1}|\mathbf{y_2},\boldsymbol\mu,\boldsymbol\Sigma \sim N_q(\boldsymbol\mu_{cond},\boldsymbol\Sigma_{cond})$$
$$\boldsymbol\mu_{cond} = \boldsymbol\mu_1 + \boldsymbol\Sigma_{12}\boldsymbol\Sigma_{22}^{-1}(\mathbf{y}_2 - \boldsymbol\mu_2), \ \ \ \ \ \ \boldsymbol\Sigma_{cond} = \boldsymbol\Sigma_{11} - \boldsymbol\Sigma_{12}\boldsymbol\Sigma_{22}^{-1}\boldsymbol\Sigma_{21}$$

The blockwise inversion formula was used to invert $\boldsymbol\Sigma$ according to the following reparameterizations.
\begin{align*}
\boldsymbol\Sigma_{11}^* & = \boldsymbol\Sigma_{11}^{-1} + \boldsymbol\Sigma_{11}^{-1}\boldsymbol\Sigma_{12}(\boldsymbol\Sigma_{22} - \boldsymbol\Sigma_{21}\boldsymbol\Sigma_{11}^{-1}\boldsymbol\Sigma_{12})^{-1}\boldsymbol\Sigma_{21}\boldsymbol\Sigma_{11}^{-1}\\
\boldsymbol\Sigma_{12}^* & = -\boldsymbol\Sigma_{11} \boldsymbol\Sigma_{12}(\boldsymbol\Sigma_{22}-\boldsymbol\Sigma_{21}\boldsymbol\Sigma_{11}^{-1}\boldsymbol\Sigma_{12})^{-1}\\
\boldsymbol\Sigma_{21}^* & = -(\boldsymbol\Sigma_{22} - \boldsymbol\Sigma_{21}\boldsymbol\Sigma_{11}^{-1}\boldsymbol\Sigma_{12})^{-1}\boldsymbol\Sigma_{21}\boldsymbol\Sigma_{11}^{-1}\\
\boldsymbol\Sigma_{22}^* & = (\boldsymbol\Sigma_{22} - \boldsymbol\Sigma_{21}\boldsymbol\Sigma_{11}^{-1}\boldsymbol\Sigma_{12})^{-1}     
\end{align*}

@@
